Towards a Sound and Complete Library for Modal Logics Reasoning in Coq

### 2022-06-19 paper/paper.tex
More writing

\begin{itemize}
  \item We define a \emph{deep embedding} of modal logic syntax in Coq and
    define its Kripke semantics.
  \item We model the Hilbert deductive systems as a Coq inductive type that is
    parameterized by different axiom systems like K, D, B, T, K4, K5, S4 and S5
    and prove its soundness with respect to Kripke semantics. We also
    discuss the current status of the formalization of the completeness result.
  \item We show a proof of the L\"ob's theorem using the Hilbert deductive system
    developed in our library.
\end{itemize}

@@

\paragraph{An overview of modal logic.} Modal logics are designed for reasoning
about truth across various --- abstract --- worlds. In such logics, a
proposition may be true in some world, but false in another one. The versions of
modal logics considered in this work extends traditional propositional logics
with two operators on propositions: $\Box$ (box) and $\Diamond$ (diamond).
The syntax of modal logic formulas are defined by the following context free
grammar (where $\circ \in\{\land,\lor,\to,\leftrightarrow\}$ and
$\mathcal{P} = \{A,B,C,...\}$ is the enumerable set of propositional variables):
\[
  \begin{array}{lcl}
    \varphi & ::= & \bot\,|\,\top\,|\,A\,|\,\neg\varphi\,|\,\varphi\circ\varphi\,|\,\Box\varphi\,|\,\Diamond\varphi
  \end{array}  
\]
Intuitively, the semantics of the operator $\Box$ is similar to an universal
quantifier, since it express \emph{necessity}: $\Box\,\varphi$ is true at the
current world iff $\varphi$ is true at all worlds. Similarly, the semantics of operator
$\Diamond$ is that a formula $\Diamond\varphi$ is true at the current
world iff $\varphi$ is \emph{possible}, i.e. true in at least some world.

@@

\begin{figure}[H]
\begin{enumerate}
  \item $\mathcal{M}\,;\,w\not\models \bot$ and $\mathcal{M}\,;\,w\models \top$;
  \item $\mathcal{M}\,;\,w\models A$ iff $\mathcal{V}(A,w) = T$;
  \item $\mathcal{M}\,;\,w\models \neg\varphi$ iff $\mathcal{M}\,;\,w\not\models\varphi$;
  \item $\mathcal{M}\,;\,w\models \varphi_1 \land \varphi_2$ iff
    $\mathcal{M}\,;\,w\models\varphi_1$ and $\mathcal{M}\,;\,w\models\varphi_2$;
  \item $\mathcal{M}\,;\,w\models \varphi_1 \lor \varphi_2$ iff
    $\mathcal{M}\,;\,w\models\varphi_1$ or $\mathcal{M}\,;\,w\models\varphi_2$;
  \item $\mathcal{M}\,;\,w\models \varphi_1 \to \varphi_2$ iff
    $\mathcal{M}\,;\,w\not\models\varphi_1$ or $\mathcal{M}\,;\,w\models\varphi_2$;
  \item $\mathcal{M}\,;\,w\models\Box\varphi$ iff $\forall w \in W. w R y \to \mathcal{M}\,;\,y \models\varphi$;
  \item $\mathcal{M}\,;\,w\models\Diamond\varphi$ iff $\exists w \in W. w R y \land \mathcal{M}\,;\,y \models\varphi$;
  \end{enumerate}
  \caption{Kripke semantics}
  \label{fig:kripke}
\end{figure}
We say that a formula $\varphi$ is satisfiable in a model $\mathcal{M}$,
denoted by $\mathcal{M}\models \varphi$, if $\forall w \in W. M\,;w\,\models \varphi$ holds.
Similarly, $\varphi$ is valid in a frame $\mathcal{F}$, written $\mathcal{F}\models\varphi$, if
it holds on all models $\mathcal{M} = \langle \mathcal{F},\mathcal{V}\rangle$, i.e.
for any $\mathcal{M}$, we have that $\mathcal{M}\models\varphi$. A formula
is valid, written $\models \varphi$, if it is valid in any frame $\mathcal{F}$.

@@

In this work, we consider Hilbert deductive systems for modal logics.
Essentially, the considered system adds axioms for the modalities
to the set of axioms for classical propositional logic.
As examples of modalities axioms we mention \textbf{K}, which express that $\Box$ distributes
over implication, and possibility which states that $\Diamond$ distributes
over disjunction.
\[
  \begin{array}{ll}
    \textbf{K}           & \Box(A \to B) \to (\Box A \to \Box B)\\
    \textbf{Possibility} & \Diamond (A \lor B) \to (\Diamond A \lor \Diamond B)\\
  \end{array}
\]
Apart from the axioms, we also need the traditional \emph{modus ponens} rule
and a rule that allows to conclude formulas with necessity.

@@

In this section, we describe the implementation details of our modal logic
library in Coq. We start by describing the structure of formula syntax and the
representation of frames and models, needed to define the Kripke semantics.
Next, we define a modular Hilbert-style deductive system. The modularization is
obtained by parameterizing it by a set of axioms used. This allow us to
generalize results about derivations to several deductive systems, independently
of the axioms used. We conclude by discussing the soundness, the current
status of the completeness and the frame correspondence proofs.

@@

\paragraph{Systems of modal logics} In order to use the deductive system to
check the provability of formulas, we need to specify the axioms that are
included in the system. Our approach is we represent to define predicates
which denote each system. The simplest set of axioms is usually refereed as
\textbf{K} and it includes the axiom $\Box(A \to B) \to (\Box A \to \Box B)$
and all the axioms for propositional logic. We represent system \textbf{K}
by the following inductive type:
\begin{minted}{coq}
Inductive K: axiom -> Prop :=
  | K_ax1: forall p q, K (ax1 p q)
  | K_ax2: forall p q r, K (ax2 p q r)
  (** some code omitted *)
  | K_axK: forall p q, K (axK p q)
\end{minted}
Type \mintinline{coq}{K} is a predicate over axioms: it has one constructor for
each of the valid axioms which form the system. Given a value
of \mintinline{coq}{axiom}, a value of type \mintinline{coq}{K a} denotes a
proof that a given axiom is part of the system \textbf{K}. This is the main
trick used by the constructor \mintinline{coq}{Ax}, of type
\mintinline{coq}{deduction}, to generically represent axioms for
different systems of modal logics uniformly. In order to understand how it
works, let us review its type (below, remeber that \mintinline{coq}{A : axiom -> Prop}
denotes a system of axioms for modal logics):
\begin{minted}{coq}
Ax: forall (t: theory)
           (a: axiom)
           (f: form),
           A a -> instantiate a = f -> deduction A t f
\end{minted}
Notice that the first parameter of constructor \mintinline{coq}{Ax} is a value
of type \mintinline{coq}{A a}, which is only inhabited if the predicate
\mintinline{coq}{A} holds for the axiom \mintinline{coq}{a}. Through this
evidence, is possible to create the corresponding axiom formula instance by
using function \mintinline{coq}{instantiate}. In this way, we are able to
represent all systems of axioms for modal logics using a single constructor
for its axioms.

\paragraph{Properties}

@@

@@

\vspace{0.5cm} {
  \begin{table}[H]
    \begin{tabular}{|c|c|c|c|c|c|c|c|c|}
      \hline
      \multirow{2}{*}{Authors} & \multicolumn{8}{c|}{Systems considered}         \\ \cline{2-9} 
                               & K & D & B & T & K4 & K5 & S4 & S5 \\ \hline
      Wind et. al.~\cite{dewind}     & X &   &   & X & X  & X  & X  &  X \\ \hline
      Doczkal et. al.~\cite{saarland}   & X &   &   & X & X  &    &    &    \\ \hline
      Benzmüller et. al.~\cite{Chris_Bruno}& X &   & X & X &  X &    &    & X  \\ \hline
      This work         & X & X & X & X & X  &  X &  X & X  \\ \hline
    \end{tabular}
    \centering
    \caption{Comparision with different Coq modal logics developements.}
    \label{tab:comparacao_trabalhos_rel}
  \end{table}
}

>>>>

\begin{itemize}
  \item We define a \emph{deep embedding} of modal logic syntax in Coq and
    define its Kripke semantics.
  \item We model the Hilbert deductive systems as a Coq inductive type that is
    parameterized by different axiom systems like K, D, B, T, K4, K5, S4, S5 and GL
    and prove its soundness with respect to Kripke semantics. We also
    discuss the current status of the formalization of the completeness result.
  \item We show a proof of the L\"ob's theorem using the Hilbert deductive system
    developed in our library.
\end{itemize}

@@

\paragraph{An overview of modal logic.} Modal logics are designed for reasoning
about truth across various --- abstract --- worlds. In such logics, a
proposition may be true in some world, but false in another one. The versions of
modal logics considered in this work extends traditional propositional logics
with two operators on propositions: $\Box$ (box) and $\Diamond$ (diamond).
The syntax of modal logic formulas are defined by the following context free
grammar (where $\circ \in\{\land,\lor,\to,\leftrightarrow\}$ and
$\mathcal{P} = \{A,B,C,...\}$ is the enumerable set of propositional variables):
\[
  \begin{array}{lcl}
    \varphi & ::= & A\,|\,\neg\varphi\,|\,\varphi\circ\varphi\,|\,\Box\varphi\,|\,\Diamond\varphi
  \end{array}  
\]
Intuitively, the semantics of the operator $\Box$ is similar to an universal
quantifier, since it express \emph{necessity}: $\Box\,\varphi$ is true at the
current world iff $\varphi$ is true at all worlds accessible from the current
world. Similarly, the semantics of operator
$\Diamond$ is that a formula $\Diamond\varphi$ is true at the current
world iff $\varphi$ is \emph{possible}, i.e. true in at least some world
accessible from the current world.

@@

\begin{figure}[H]
\begin{enumerate}
  \item $\mathcal{M}\,;\,w\not\models \bot$ and $\mathcal{M}\,;\,w\models \top$;
  \item $\mathcal{M}\,;\,w\models A$ iff $\mathcal{V}(A,w) = T$;
  \item $\mathcal{M}\,;\,w\models \neg\varphi$ iff $\mathcal{M}\,;\,w\not\models\varphi$;
  \item $\mathcal{M}\,;\,w\models \varphi_1 \land \varphi_2$ iff
    $\mathcal{M}\,;\,w\models\varphi_1$ and $\mathcal{M}\,;\,w\models\varphi_2$;
  \item $\mathcal{M}\,;\,w\models \varphi_1 \lor \varphi_2$ iff
    $\mathcal{M}\,;\,w\models\varphi_1$ or $\mathcal{M}\,;\,w\models\varphi_2$;
  \item $\mathcal{M}\,;\,w\models \varphi_1 \to \varphi_2$ iff
    $\mathcal{M}\,;\,w\not\models\varphi_1$ or $\mathcal{M}\,;\,w\models\varphi_2$;
  \item $\mathcal{M}\,;\,w\models\Box\varphi$ iff $\forall y \in W. w R y \to \mathcal{M}\,;\,y \models\varphi$;
  \item $\mathcal{M}\,;\,w\models\Diamond\varphi$ iff $\exists y \in W. w R y \land \mathcal{M}\,;\,y \models\varphi$;
  \end{enumerate}
  \caption{Kripke semantics}
  \label{fig:kripke}
\end{figure}
We say that a formula $\varphi$ is satisfiable in a model $\mathcal{M}$,
denoted by $\mathcal{M}\models \varphi$, if $\forall w \in W. M\,;w\,\models \varphi$ holds.
Similarly, $\varphi$ is valid in a frame $\mathcal{F}$, written $\mathcal{F}\models\varphi$, if
it holds on all models $\mathcal{M} = \langle \mathcal{F},\mathcal{V}\rangle$, i.e.
for any $\mathcal{M}$, we have that $\mathcal{M}\models\varphi$. A formula
is valid, written $\models \varphi$, if it is valid in any frame $\mathcal{F}$.

@@

In this work, we consider Hilbert deductive systems for modal logics.
Essentially, the considered system adds axioms for the modalities
to the set of axioms for classical propositional logic.
As examples of modalities axioms we mention \textbf{K}, which express that $\Box$ distributes
over implication, and possibility which states that $\Diamond$ distributes
over disjunction.
\[
  \begin{array}{ll}
    \textbf{K} & \Box(\varphi_1 \to \varphi_2) \to (\Box \varphi_1 \to \Box \varphi_2)\\
    \textbf{Possibility} & \Diamond (\varphi_1 \lor \varphi_2) \to
                           (\Diamond \varphi_2 \lor \Diamond \varphi_2)\\
  \end{array}
\]
Apart from the axioms, we also need the traditional \emph{modus ponens} rule
and a rule that allows to conclude formulas with necessity.

@@

In this section, we describe the implementation details of our modal logic
library in Coq. We start by describing the structure of formula syntax and the
representation of frames and models, needed to define the Kripke semantics.
Next, we define a modular Hilbert-style deductive system. The modularization is
obtained by parameterizing it by a set of axioms used. This allows us to
generalize results about derivations to several deductive systems, independently
of the axioms used. We conclude by discussing the soundness, the current
status of the completeness and the frame correspondence proofs.

@@

\paragraph{Systems of modal logics} In order to use the deductive system to
check the provability of formulas, we need to specify the axioms that are
included in the system. Our approach is to define predicates
which denote each system. The simplest set of axioms is usually refereed as
\textbf{K} and it includes the axiom $\Box(A \to B) \to (\Box A \to \Box B)$
and all the axioms for propositional logic. We represent system \textbf{K}
by the following inductive type:
\begin{minted}{coq}
Inductive K: axiom -> Prop :=
  | K_ax1: forall p q, K (ax1 p q)
  | K_ax2: forall p q r, K (ax2 p q r)
  (** some code omitted *)
  | K_axK: forall p q, K (axK p q)
\end{minted}
Type \mintinline{coq}{K} is a predicate over axioms: it has one constructor for
each of the valid axioms which form the system. Given a value
of \mintinline{coq}{axiom}, a value of type \mintinline{coq}{K a} denotes a
proof that a given axiom is part of the system \textbf{K}. This is the main
trick used by the constructor \mintinline{coq}{Ax}, of type
\mintinline{coq}{deduction}, to generically represent axioms for
different systems of modal logics uniformly. In order to understand how it
works, let us review its type (below, remeber that \mintinline{coq}{A : axiom -> Prop}
denotes a system of axioms for modal logics):
\begin{minted}{coq}
Ax: forall (t: theory)
           (a: axiom)
           (f: form),
           A a -> instantiate a = f -> deduction A t f
\end{minted}
Notice that the first parameter of constructor \mintinline{coq}{Ax} is a value
of type \mintinline{coq}{A a}, which is only inhabited if the predicate
\mintinline{coq}{A} holds for the axiom \mintinline{coq}{a}. Through this
evidence, is possible to create the corresponding axiom formula instance by
using function \mintinline{coq}{instantiate}. In this way, we are able to
represent all systems of axioms for modal logics using a single constructor
for its axioms.

\paragraph{Properties} A key property of the Hilbert deductive system formalized
in our library is its soundness. If a formula $\varphi$ is derived from a
context $\Gamma$, i.e. $\Gamma\vdash\varphi$, then $\varphi$ is entailed from
the set $\Gamma$, i.e. $\Gamma\models\varphi$. The Coq statement of the soundness
result is:

\begin{minted}{coq}
Theorem soundness:
  forall G p, (K ; G |-- p) -> (G ||= p).
\end{minted}

The proof proceeds by induction on the derivation of $\Gamma \vdash\varphi$ and
it has a case for each of the 11 axioms of the Hilbert system for propositional
logics, one for the $K$ axiom and cases for using hypothesis and the rules modus
ponens. In order to modularize the proof, we defined a lemma for each case.
As example, the following is the statement for modus ponens case.
\begin{minted}{coq}
Lemma Modus_Ponens_soundness:
  forall M w p q,
  ((M ' w ||- p) /\
  (M ' w ||- [! p -> q !])) ->
  (M ' w ||- q).
\end{minted}
All these lemmas are direct consequences from the definitions of the Kripke
semantics and entailment.

While we have completely proved the soundness theorem, the current version of
our library does not have a completeness theorem.

@@

In this section we describe a case study of the defined modal logic library.
We proved the L\"ob theorem in the system K4 with modal fixed points.
The importance of L\"ob's theorem is due to its use in provability
logic~\cite{boolos_1994} to abstract details from encodings in G\"odel's
incomleteness results.

@@

\vspace{0.5cm} {
  \begin{table}[H]
    \begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}
      \hline
      \multirow{2}{*}{Authors} & \multicolumn{9}{c|}{Systems considered} \\ \cline{2-10} 
                               & K & D & B & T & K4 & K5 & S4 & S5 & GL\\ \hline
      Wind et. al.~\cite{dewind}     & X &   &   & X & X  & X  & X  &  X & \\ \hline
      Doczkal et. al.~\cite{saarland}   & X &   &   & X & X  &    &    &  &  \\ \hline
      Benzmüller et. al.~\cite{Chris_Bruno}& X &   & X & X &  X &    &    & X &  \\ \hline
      This work         & X & X & X & X & X  &  X &  X & X  & X \\ \hline
    \end{tabular}
    \centering
    \caption{Comparision with different Coq modal logics developements.}
    \label{tab:comparacao_trabalhos_rel}
  \end{table}
}
